---
title: "Is an Automatic or Manual Transmission Better for MPG?"
author: "Matthew Sedlar"
date: "October 19, 2015"
output: html_document
---

## Executive Summary

## Exploratory Analysis

I started by looking at simple box plot of miles per gallon by automatic and manual transmission types. As figure 1 shows, the interquartile range for automatic transmissions is below the interquartile range of manual transmissions. There certainly appears to be a relationship better mpg and manual transmissions.

```{r echo=FALSE, message=FALSE}

library(ggplot2)
library(gridExtra)

# calling up and cloning data set for analysis
# converting am variable to factor
data(mtcars)
analysis <- mtcars
analysis$am <- sapply(analysis$am, function(x) ifelse(x==0, "automatic","manual"))

p <- ggplot(analysis) + geom_boxplot(aes(x=am,y=mpg)) +
  xlab("Transmission Type") +
  ylab("Miles per Gallon (U.S.)") +
  ggtitle("Miles per Gallon by Transmission Type")
p

```

While the difference between the average mpg for the transmission types is statistically significant (see Appendix A), a simple categorical variable cannot fully explain _why_ there is a difference. If you were to run out and buy any car with a manual transmission, what other variables result in better miles/gallon? We have to explore other variables.



## Testing Independent X Variables

## Model

## Hypothesis Testing

## Conclusion

---

# Appendix

### A: Automatic vs. Manual Hypothesis Testing

$$H_0: \bar{X} = \bar{Y}$$
$$H_1: \bar{X} \neq \bar{Y}$$

```{r}
transa <- subset(analysis,am=="automatic") 
transm <- subset(analysis,am=="manual")

t.test(transa$mpg,transm$mpg)
```

### B: Independent Variables by MPG, Faceted by Transmission Type

```{r echo=FALSE}
p.wt <- ggplot(analysis) + geom_point(aes(x=wt,y=mpg)) + facet_grid(.~am) +
  xlab("weight")

p.cyl <- ggplot(analysis) + geom_point(aes(x=cyl,y=mpg)) + facet_grid(.~am) +
  xlab("number of cylinders")

p.hp <- ggplot(analysis) + geom_point(aes(x=hp,y=mpg)) + facet_grid(.~am) +
  xlab("horsepower")

p.disp <- ggplot(analysis) + geom_point(aes(x=disp,y=mpg)) + facet_grid(.~am) +
  xlab("displacement")

p.carb <- ggplot(analysis) + geom_point(aes(x=carb,y=mpg)) + facet_grid(.~am) +
  xlab("number of carburetors")

grid.arrange(p.wt,p.cyl,p.hp, p.disp,p.carb,ncol=3,nrow=2)
```